---
title: "Hierarchical clustering"
author: "Jindra"
date: "12/1/2017"
output: html_document
---

### Setup
```{r setup}
suppressMessages(library(dplyr))
library(tidyr)
library(ggplot2)
library(mlbench)
pca <- iris %>%
  select(-Species) %>%
  prcomp

mapped_iris <- pca %>%
  predict(iris)
```

#Hierarchichal clustering
Hierarchical clustering is a technique you can use when you have a distance matrix of your data. Here the idea is that you build up a tree structure of nested clusters by iteratively merging clusters. You start with putting each data point in their own singleton clusters. Then iteratively you find two clusters that are close together and merge them into a new cluster. You continue this until all data points are in the same large cluster.
```{r hierarchical_clustering}
iris_dist <- iris %>% select(-Species) %>% scale %>% dist
clustering <- hclust(iris_dist)
plot(clustering) #generic plot

#install.packages("ggdendro")
library(ggdendro)
ggdendrogram(clustering) + theme_dendro()
```

Only visualizing the clustering is rarely enough, so to work with the result we need to be able to extract the actual groupings. Since we are working with the iris data, it is natural to want to split the data into three clusters.
```{r}
clusters <- clustering %>% cutree(k = 3)
iris %>%
  cbind(Cluster = clusters) %>%
  ggplot() +
  geom_bar(aes(x = Species, fill = as.factor(Cluster)),
           position = "dodge") +
  scale_fill_discrete("Cluster")

#plot
mapped_iris %>%
  as.data.frame %>%
  cbind(Species = iris$Species,
        Clusters = as.factor(clusters)) %>%
  ggplot() +
  geom_point(aes(x = PC1, y = PC2,
                   shape = Species, colour = Clusters))
```

