---
title: "tidyr package"
author: "Jindra"
date: "11/20/2017"
output: html_document
---

## Tidying the data with tidyr
```{r}
#install.packages("tidyr")
library(tidyr)
library(dplyr)
library(ggplot2)
#tidy data mean different things, for example easy summarizing and plotting
iris %>% select(Species, Petal.Length) %>% head(3)
iris %>% select(Species, Petal.Length) %>%
  qplot(Species, Petal.Length, geom = "boxplot", data = .)
```
But what happens if I want to plot the different measurements of the irises to see how those are? Each measurement is a separate column. They are Petal.Length, Petal.Width, and so on.

Now I have a bit of a problem because the different measurements are in different columns in my data frame. I cannot easily map them to an x-axis and a y-axis.

The tidyr package addresses that

### gather
It modifies the data frame so columns become names in a factor and other columns become values
What it does is essentially transform the data frame so that you get one column containing the name of your original columns and another column containing the values in those columns
```{r}
iris %>%
  gather(key = Attribute, value = Measurement, Sepal.Length, Sepal.Width) %>%
  select(Species, Attribute, Measurement) %>%
  head(3)

#easy plotting of combined attributes
iris %>%
  gather(key = Attribute, value = Measurement, Sepal.Length, Sepal.Width) %>%
  select(Species, Attribute, Measurement) %>%
  qplot(Attribute, Measurement, geom = "boxplot", facets = . ~ Species, data = .)
```

