---
title: "Expressions revisited"
author: "Jindra"
date: "12/3/2017"
output: html_document
---
## Expressions and precedence
```{r precedence}
x = 5
+x
-x

#Division will return a floating-point number even if both its operands are integers
10/x
11/x

#integer division
9 %/% x

#remainder of integer division
9 %% x

#exponentiation
x^2
x**2

#precedence of ":" operator - exonentiation -> ":" -> multiplication
1:2**2 #exponentiation first
1:2*2 # ":" first
-1:2 # "-" first
-(1:2)

#functions are evaluated before the operators:
1:sqrt(4)
```

### Boolean expressions
```{r bool}
TRUE
T
FALSE
F

#negate the expression
!TRUE  # = FALSE
!FALSE # = TRUE

#The | and || are logical “or” operators while & and && are logical “and” operators.
T | F
F || F
T & F
T && T
```

The difference between | and || or & and && are how they deal with vectors. The one-character version will apply the operator element-wise and create a vector while the two-character version will only look at the first value in vectors.
```{r bool_2}
x <- c(TRUE, FALSE, TRUE, FALSE)
y <- c(TRUE, TRUE, FALSE, FALSE)

x | y
x || y
x & y
x && y
```

We typically use the two-character version in control structures like if—since these do not operate on vectors in any case. We use the one-character version when we need to compute with Boolean arithmetic, when we want our expressions to work as vectorized expressions.  
Incidentally, all the arithmetic operators work like the | and & operators when operating on more than one value, i.e., they operate element-wise on vectors.

## Basic data types
There are a few basic types in R: numeric, integer, complex, logical, and character.
```{r types}
#The numeric type is what you get any time you write a number into R
is.numeric(2)
class(2)
```

The integer type is used for, well, integers. Surprisingly, the 2 is not an integer in R. It is a numeric type which is the larger type that contains all floating-point numbers as well as integers. To get an integer you have to make the value explicitly an integer, and you can do that using the function as.integer or writing L after the literal.
```{r types_2}
is.integer(2)
is.integer(2L)
x <- as.integer(2)
is.integer(x)
class(x)
x <- c(2L, 3L, 4L) #integers
x <- c(2, 3, 4) #numerics

#If you translate a non-integer into an integer, you just get the integer part.
as.integer(3.3)
as.integer(9.9)
```

### Complex numbers
```{r complex_types}
1 + 0i
is.complex(1 + 0i)
class(1 + 0i)
sqrt(as.complex(-1))
```

### Logical types
```{r logic}
x <- 5 > 4
x
class(x)
is.logical(x)
```

### Character type
```{r}
x <- "hello, world"
class(x)
is.character(x)

#Convertions between types:
as.character(3.14)
```

Unlike in some languages, character doesn’t mean a single character but any text. So it is not like in C or Java where you have single character types, 'c', and multi-character strings, "string", they are both just characters.
