---
title: "Switching to C++"
author: "Jindra"
date: "1/27/2018"
output: html_document
---

This is a drastic step, but by switching to a language such as C++, you have more fine-grained control over the computer. Of course, this also means that you don’t have many of the features that R does, so you don’t want to program an entire analysis in C++, but you might want to translate the time-critical code to C++.

### Rcpp package
Luckily, the Rcpp package makes integrating R and C++ pretty straightforward. The only thing to really be careful about is that C++ index from 0 and R from 1. Rcpp takes care of converting this so a 1 indexed vector from R can be accessed as a 0 indexed vector in C++, but when translating code you have to keep it in mind.  
When we have a relatively simple function that is not using any of R’s advanced features it is ideal to translate into C++.  
```{r}
#install.packages("Rcpp")
library(Rcpp)
cppFunction("
NumericVector smooth_weights_iteration_cpp(List g,
                             NumericVector node_weights,
                             double alpha)
{
  NumericVector new_weights(g.length());

  for (int i = 0; i < g.length(); ++i) {

    IntegerVector neighbours = g[i];
    double neighbour_sum = 0.0;
    int n = 0;

    for (int j = 0; j < neighbours.length(); ++j) {
      neighbour_sum += node_weights[j];
      ++n;
    }

    if (n > 0) {
      new_weights[i] = alpha * node_weights[i] +
        (1-alpha) * (neighbour_sum / n);
    } else {
      new_weights[i] = node_weights[i];
    }
  }

  return new_weights;
}
")
```

This code creates a function with the same name as the C++ function that can be called directly from R and Rcpp will take care of converting types as needed.
```{r}
smooth_weights_cpp <- function(graph, node_weights,
                               alpha, no_iterations) {
  new_weights <- node_weights
  for (i in 1:no_iterations) {
    new_weights <- smooth_weights_iteration_cpp(graph, new_weights, alpha)
  }
  new_weights
}
```

If we compare the R and C++ function, we see that we get a substantial performance boost from this.  
```{r}
library(microbenchmark)
microbenchmark(
  smooth_weights(g, weights, 0.8, 10),
  smooth_weights_cpp(g, weights, 0.8, 10),
  times = 5
)
```


If you have a few performance hotspots in your code that are relatively simple, just very time-consuming because they do a lot of work, it is worth considering translating these to C++ and Rcpp makes it easy.  
Don’t go overboard, though. It is harder to profile and debug code in C++, and it is harder to refactor your code if it is a mix of C++ and R. Use it, but use it only when you really need it.
