---
title: "Reading Data"
author: "Jindra"
date: "11/19/2017"
output: html_document
---

### Data Reading
If the data format is a standard table format, use built-in functions
```{r}
?read.table
```

### Reading and formatting datasets
```{r}
library(mlbench)
data(BreastCancer)

library(magrittr)
BreastCancer %>% head(3)
```

### Read data from a URL
```{r}
data_url = "http://tinyurl.com/kw4xtts"
lines = readLines(data_url)
lines[1:5]
```

### Save the data to a file (target folder needs to be created first)
```{r}
#make sure that your working directory is set properly
getwd()
# if necessary - setwd()
writeLines(lines, con = "data/raw-breast-cancer.csv") #you can use tab completion
```

### Read the created csv file
```{r}
raw_breast_cancer <- read.csv("data/raw-breast-cancer.csv")
raw_breast_cancer %>% head(3)
```

The data can be of course read straigh from the URL, since it is a csv in first place
```{r}
raw_breast_cancer_url <- read.csv(data_url)
raw_breast_cancer_url %>% head(4)
```

First line in BreastCancer seems to have been turned into column names, fix it with header parameter
```{r}
raw_breast_cancer <- read.csv("data/raw-breast-cancer.csv", header = FALSE)
raw_breast_cancer %>% head(3)
```

Now the first line is not used for the header, but we still don't know what are the columns
Normally, we would find metadata, here we will cheat and use the dataset from mlbench
```{r}
names(raw_breast_cancer) = names(BreastCancer)
raw_breast_cancer %>% head(3)
#it can be set also during loading
#raw_breast_cancer <- read.csv(data_url, header = FALSE, col.names = names(BreastCancer))
```


### More data cleaning
The Class column is not right - we have numbers (2 and 4), but it means (benign and malign) - let's copy the data, since we don't want to modify original data, and change it
```{r}
formatted_breast_cancer = raw_breast_cancer
formatted_breast_cancer$Class %>% unique
#2 and 4 are the only values
map_class <- function(x){
  ifelse(x == 2,'benign','malign')
  #the result does not need to be assigned to a variable, if it is the last value
  #x = ifelse(x == 2,'benign','malign')
  #the safer way of doing this would be inlcusion of possible NAs, if more possibilies would be in data accidentaly
  #ifelse(x == 2, 'benign',ifelse(x == 4, 'malign', NA))
}
mapped = formatted_breast_cancer$Class %>% map_class
mapped %>% head(6)
mapped %>% table
```

Other option of data cleaning would be dictionaries
```{r}
dict = c("2" = "benign", "4" = "malign")
#new map funciton
map_class = function(x) dict[as.character(x)]
mapped = formatted_breast_cancer$Class %>% map_class
mapped %>% table
#the vector looks a little strange though...
mapped[1:5]
#This is because when we create a vector by mapping in this way we preserve the names of the values.
mapped %<>% unname
mapped[1:5]
```

Now we just need to translate this vector of strings into a factor
```{r}
formatted_breast_cancer$Class = mapped %>% factor(levels = c("benign","malign"))
formatted_breast_cancer %>% head(3)
```

### The whole process in one chunk
```{r}
read.csv(data_url, header = FALSE,
         col.names = names(BreastCancer)) ->
  raw_breast_cancer ->
  formatted_breast_cancer

dict <- c("2" = "benign", "4" = "malignant")
map_class <- function(x) dict[as.character(x)]
formatted_breast_cancer$Class <-
  formatted_breast_cancer$Class %>%
  map_class %>%
  unname %>%
  factor(levels = c("benign", "malignant"))

#it can be also saved - .rda stands for R data
formatted_breast_cancer %>% save(file = "data/formatted-breast-cancer.rda")

#the load function gets even the original variable name, so it is not needed to assign the result to a new variable
#load("data/formatted-breast-cancer.rda")
```

## With anonymous function - if we don't want to name map_class
```{r}
raw_breast_cancer$Class %>%
  { dict <- c("2" = "benign", "4" = "malignant")
    dict[as.character(.)]
  } %>%
  unname %>%
  factor(levels = c("benign", "malignant")) %>%
  table
```
