---
title: "Linear regression"
author: "Jindra"
date: "11/26/2017"
output: html_document
---
## Simple linear regression
If we take a simple linear regression, f_θ(x) = θ_1 x + θ_0, we need the function lm().
```{r simple_linear_regression}
library(dplyr)
library(ggplot2)
data(cars)
cars %>% head

#we can see clear linear relationship from the plot
cars %>% qplot(speed,dist,data = .)
cars %>% lm(dist ~ speed, data = .)

#use ggplot to create linear regression
cars %>% ggplot(aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(expand = c(0,0), limits = c(2,26), breaks = seq(2,26,4)) +
  scale_y_continuous(expand = c(0,0), limits = c(-15,125))
```
By default geom_smooth() will also plot the uncertainty of the fit. This is the gray area in the plot. This is the area where the line is likely to be (assuming that the data is generated by a linear model). Do not confuse this with where data points are likely to be, though

### Linear regression - analysis
```{r lm_function}
cars %>% lm(dist ~ speed, data = .) %>% summary
cars %>% lm(dist ~ speed, data = .) %>% coefficients
cars %>% lm(dist ~ speed, data = .) %>% confint
```
Here, (Intercept) is θ_0 and speed is θ_1.

### Comparison of different parameters
```{r parameter_comparison}
library(magrittr)
predict_dist <- function(speed, theta_1){
  data.frame(speed = speed,
             dist = speed * theta_1,
             theta = as.factor(theta_1))
}
cars %>% ggplot(aes(x = speed, y = dist, colour = theta)) +
  geom_point(colour = "black") +
  geom_line(data = predict_dist(cars$speed,2)) +
  geom_line(data = predict_dist(cars$speed,3)) +
  geom_line(data = predict_dist(cars$speed,4)) +
  scale_color_discrete(name=expression(theta[1]))
```

### Error function
```{r error_function}
thetas = seq(0, 5, length.out = 50)
fitting_error = Vectorize(function(theta)
  sum((theta * cars$speed - cars$dist)**2)
  )
data.frame(thetas = thetas, errors = fitting_error(thetas)) %>%
  ggplot(aes(x = thetas, y = errors)) +
  geom_line() +
  #scale_x_continuous(expression(theta[1]))
  xlab(expression(theta[1])) +
  ylab(" ")
```

### Formula for zero intercept
```{r zero_intercept}
cars %>% lm(dist ~ speed - 1, data = .) %>% summary
cars %>% lm(dist ~ speed - 1, data = .) %>% coefficients
cars %>% lm(dist ~ speed, data = .) %>% coefficients

#line comparison
x_d = 0:30
zero_interc = 2.909132 * x_d
with_interc = 3.932409 * x_d - 17.579095
ggplot() +
  ggtitle("Comparison of intercept choice") +
  geom_line(aes(x = x_d, y = zero_interc, colour = "None")) +
  geom_line(aes(x = x_d, y = with_interc, colour = "With")) +
  geom_point(aes(x = cars$speed, y = cars$dist)) +
  xlab("Speed") +
  ylab("Distance") +
  scale_colour_manual("Intercept", values = c("None"="red", "With"="blue"))
```

### Zero intercept with geom_smooth()
```{r intercept_smooth}
cars %>% ggplot(aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x - 1) +
  ggtitle("Zero intercept with geom_smooth()") +
  xlab("Speed") +
  ylab("Distance")
#aes needs to be specified "globally" in ggplot() - if it is specified in geom_point(), variables x and y are unknown

```

### Binary classification
```{r binary_classification}
library(mlbench)
data("BreastCancer")
BreastCancer %>% head
```

